---
title: "ENVS 193DS Homework #3"
author: "Ella Menin"
format: pdf
---

# Setting up

```{r}
#| label: reading-in-packages
#| message: false
#| warning: false

# reading in package: tidyverse
library(tidyverse)
# reading in package: here
library(here)
# reading in package: janitor
library(janitor)
# reading in package: readxl
library(readxl)

```

```{r}
#| label: reading-in-kelp-data

# storing read kelp data as "kelp" object
kelp <- read.csv("data/temp-kelp.csv")

```

```{r}
#| label: reading-in-personal-data

# storing read personal data as "personal_data" object
personal_data <- read.csv("data/personal data(Sheet1).csv")

```

# Problem 1

## a.

The appropriate tests to determine the strength of the relationship between temperature and giant kelp frond elongation rate are a Pearson correlation and and a simple linear regression. A Pearson correlation evaluates the strength and direction of the linear relationship between two variables; in this case, ocean temperature and giant kept frond elongation rate. A simple linear regression also evaluates the relationship between two variables, but it specifically shows elongation rate as a response variable that is predicted by temperature. Creating a linear model for this data can show how much growth changes per 1 degree Celsius increase in temperature.

## b.

```{r}
#| label: lm-model-and-predicted-values

# creating new object called "kelp lm" out of linear model evaluating relationship between kept elongation rate and temperature in deg celsius
kelp_lm <- lm(kelp_elong ~ temp_c, data = kelp) # kelp data frame

# creating object kelp_preds by combining (cbind) the original kelp data set and predicted values and confidence intervals from the linear model above 
kelp_preds <- cbind(
  # original data set 
  kelp,
  # generating predicted values from linear model 
  predict(kelp_lm, interval = "confidence") # interval = "confidence" calculates lower and upper confidence intervals and predicted regression line values
)

```


```{r}
#| label: creating-visualization

# base layer ggplot with kelp_preds data frame
ggplot(data = kelp_preds, 
       # setting aesthetics 
       mapping = aes(
         # x is temperature measured in celsius
         x = temp_c,
         # y is giant kelp frond elongation rate measured in cm per day
         y = kelp_elong)) +
  # first layer geom_ribbon to visually show confidence intervals for linear model
  geom_ribbon(aes
              # setting the ribbon bounds to be the lower and upper confidence intervals
              (ymin = lwr, ymax = upr),
              # making our ribbon a nice light blue color
              fill = "lightblue",
              # making it so that you can see through our ribbon
              alpha = 0.4) +
  # second layer geom_line to show relationship line
  geom_line(aes
            # uses predicted values previously generated from lm as y-values for our line
            (y = fit),
            # color that compliments our ribbon
            color = "steelblue",
            # adjusting width of the line
            linewidth = 1.2) +
  # third layer geom_point to show individual observations in pretty green color
  geom_point(color = "darkgreen",
             # making points noticably large
             size = 3) +
  # using labs to set axis titles
  labs(
    # x axis: ocean temperature
    x = "Ocean Temperature (°C)",
    # y axis: giant kelp frond elongation rate
    y = "Giant Kelp Frond Elongation Rate (cm day$^{-1}$)",
    # visualization title: finding from visualization
    title = "As ocean temperature decreases, so does elongation rate"
  ) +
  # setting a nice, basic theme
  theme_minimal()

```

## c.

### Running tests:

```{r}

# creating new object called "kelp lm" out of linear model evaluating relationship between kept elongation rate and temperature in deg celsius (done in previous part, but wanted to show again for this problem part)
kelp_lm <- lm(kelp_elong ~ temp_c, data = kelp) # data frame kelp

# display summary results of kelp linear model
summary(kelp_lm)

```

**Note:** I decided that showing my code in which I ran my test is appropriate in this situation because this same code is used in the next part to check my assumptions.

### Checking assumptions:

```{r}
#| label: checking-assumptions

# creating diagnostic plots to check model assumptions
par(mfrow = c(2, 2))
# plot assumption plots: residuals vs fitted, q-q residuals, scale-location, and residuals vs leverage
plot(kelp_lm)

```

Here, I checked the assumptions of the linearity of data, the normality of residuals, homoscedasticity (constant variance), and if there are outliers significant enough to influence our plot.

The residuals vs fitted and scale-location plots featured points that are somewhat evenly and randomly distributed around the horizontal dotted line, indicating homoscedasticity. The Q-Q residuals plot suggests normally enough distributed residuals. Finally, the residuals vs leverage plot does not reveal any outliers significant enough to heavily influence the model, as all points fall within the indicator lines (indicator lines are not able to be seen, suggesting that points fall within them). All of these factors indicate that a linear model is appropriate for this data. 

## d.

To evaluate the relationship between ocean temperature and giant kelp frond elongation rate, a simple linear regression was used to test whether temperature could predict kelp growth.

The results suggest to reject the null hypothesis because a strong negative relationship was found between ocean temperature and giant kelp frond elongation rate, meaning kelp grew more slowly at higher temperatures (linear regression, F(1, 30) = 26.93, R² = 0.473, p < 0.001, $\alpha$ = 0.05). On average, kelp elongation rate decreased by about 0.43 cm day$^{-1}$ for every 1°C increase in ocean temperature.

## e.

These results suggest that warmer ocean temperature (°C) could reduce giant kelp frond elongation rate (cm day$^{-1}$). Because giant kelp forests provide crucial habitat and food resources for marine animals, a slower growth rate of these giant kelp fronds could result in resource availability for the dependent species. As the ongoing climate crisis worsens, ocean temperatures are projected to continue rising, which can further encourage the slowing of giant kelp frond elongation. Furthermore, it is crucial to continue investing time and funding into developing solutions to rising sea temperatures.

## f.

```{r}
#| label: pearson-correlation-test

# pearson correlation test to assess strength and direction of relationship
cor.test(
  # first variable: temperature from kelp data frame
  kelp$temp_c,
  # second variable: elongation rate from kelp data frame
  kelp$kelp_elong,
  # specification that it is a pearson correlation test
  method = "pearson"
)

```

The Pearson correlation test and the simple linear regression produced the same overall conclusion: to reject the null hypothesis in favor of accepting the alternative hypothesis, as both tests found a statistically significant negative relationship between ocean temperature and giant kelp frond elongation rate.

The Pearson correlation measured the strength and direction of the relationship between temperature and kelp elongation rate (r = −0.688), while the linear regression estimated the amount of change in kelp growth associated with temperature (kelp elongation rate decreased by approximately 0.43 cm day$^{-1}$ for every 1°C increase in ocean temperature). 