---
title: "ENVS 193DS Homework #3"
author: "Ella Menin"
format: pdf
---

# Setting up

```{r}
#| label: reading-in-packages
#| message: false
#| warning: false

# reading in package: tidyverse
library(tidyverse)
# reading in package: here
library(here)
# reading in package: janitor
library(janitor)
# reading in package: readxl
library(readxl)

```

```{r}
#| label: reading-in-kelp-data

# storing read kelp data as "kelp" object
kelp <- read.csv("data/temp-kelp.csv")

```

```{r}
#| label: reading-in-personal-data

# storing read personal data as "personal_data" object
personal_data <- read.csv("data/personal data(Sheet1).csv")

```

# Problem 1

## a.

The appropriate tests to determine the strength of the relationship between temperature and giant kelp frond elongation rate are a Pearson correlation and and a linear model. A Pearson correlation evaluates the strength and direction of the linear relationship between two variables; in this case, ocean temperature and giant kept frond elongation rate. A linear model also evaluates the relationship between two variables, but it specifically shows elongation rate as a response variable that is predicted by temperature. Creating a linear model for this data can show how much growth changes per 1 degree Celsius increase in temperature.
